One-Particle Density Matrix Polarization Susceptibility Tensors

### 2019-02-04 papers/1_dmatpol/revision/paper-revision-1.tex
Test of gamma scaling of XC density

\author{Bartosz B{\l}asiak}
\email[]{blasiak.bartosz@gmail.com}
\homepage[]{http://www.polonez.pwr.edu.pl}
\affiliation{Department of Physical and Quantum Chemistry, Faculty of Chemistry, 
Wroc{\l}aw University of Science and Technology, 
Wybrze{\.z}e Wyspia{\'n}kiego 27, Wroc{\l}aw 50-370, Poland}

@@

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals (MO's) as linear combinations
of AO's, referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} =  {\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)}  \;.
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true in a special case when the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$.
By substituting the McWeeny's result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} one can show that
%
%\begin{multline} \label{e:dmu-4-exact.linear-approximation}
% \frac{1}{2} 
% \delta {\BM{\upmu}}({\bf r}_Q)
%  \cong
%   \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
%    \right] }
%%
%  +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
%    \right] } \\
%%
%  -\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
%    \right] }
%%
%  -\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
%    \right] } \;,
%\end{multline}
%%
%which can be recast to
%
\begin{equation}\label{e:dmu-l-vector}
 \frac{1}{4} 
 \delta {\BM{\upmu}}({\bf r}_Q)
  \cong
  \Re \left\{ 
 \Tr{ 
    \left[ 
         \delta {\bf C} \cdot \mathbb{L}({\bf r}_Q)
   \right] }  
  \right\} \;.
\end{equation}
%
In the above expression, the `$\Re$' symbol denotes the real part,
and the intermediate quantity
%
\begin{equation} \label{e:L-projector}
 \mathbb{L}({\bf r}_Q) \equiv {\bf C}^{(0)\dagger} \cdot \mathbb{M}({\bf r}_Q) \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \;.
\end{equation}
%
Note that $\mathbb{L}({\bf r}_Q)$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
%Note also the similarity between Eq.~\eqref{e:L-projector} and Eq.~\eqref{e:dmu-exact}.

@@

Once the approximate relationship between the orbital change and the external electric field
has been established in Eq.~\eqref{e:mu-ind.matrix.linear.solution},
one can obtain the final expression for the density matrix change 
that reproduces the induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
%\begin{multline}  \label{e:susceptibility-B} % PREPRINT
\begin{equation}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i}^{(0)} {\bf b}_{\beta}^{(i;1)} + C_{\beta i}^{(0)} {\bf b}_\alpha^{(i;1)} %\\ PREPRINT
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i}^{(0)} + D_{\beta\gamma} C_{\alpha i}^{(0)}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{equation}
%\end{multline} % PREPRINT
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ 
describes the local linear polarization affinity 
of the density matrix element 
associated with the $i$th LMO centroid as well as the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all the elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule. 
However, they are valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavor of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. \InlineImage{eu-logo.eps} This project is funded by National Science Centre, Poland 
(registration No.~2016/23/P/ST4/01720) within the POLONEZ 3 fellowship. 
\end{acknowledgments}

>>>>

\author{Bartosz B{\l}asiak}
\email[]{blasiak.bartosz@gmail.com}
\homepage[]{http://www.polonez.pwr.edu.pl}
\affiliation{Department of Physical and Quantum Chemistry, Faculty of Chemistry, 
Wroc{\l}aw University of Science and Technology, 
Wybrze{\.z}e Wyspia{\'n}skiego 27, Wroc{\l}aw 50-370, Poland}

@@

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals (MO's) as linear combinations
of AO's, referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)}  \;.
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true in a special case when the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$.
By substituting the McWeeny's result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} one can show that
%
%\begin{multline} \label{e:dmu-4-exact.linear-approximation}
% \frac{1}{2} 
% \delta {\BM{\upmu}}({\bf r}_Q)
%  \cong
%   \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
%    \right] }
%%
%  +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
%    \right] } \\
%%
%  -\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
%    \right] }
%%
%  -\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
%    \right] } \;,
%\end{multline}
%%
%which can be recast to
%
\begin{equation}\label{e:dmu-l-vector}
 \frac{1}{4} 
 \delta {\BM{\upmu}}({\bf r}_Q)
  \cong
  \Re \left\{ 
 \Tr{ 
    \left[ 
         \delta {\bf C} \cdot \mathbb{L}({\bf r}_Q)
   \right] }  
  \right\} \;.
\end{equation}
%
In the above expression, the `$\Re$' symbol denotes the real part,
and the intermediate quantity
%
\begin{equation} \label{e:L-projector}
 \mathbb{L}({\bf r}_Q) \equiv {\bf C}^{(0)\dagger} \cdot \mathbb{M}({\bf r}_Q) \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \;.
\end{equation}
%
Note that $\mathbb{L}({\bf r}_Q)$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
%Note also the similarity between Eq.~\eqref{e:L-projector} and Eq.~\eqref{e:dmu-exact}.

@@

Once the approximate relationship between the orbital change and the external electric field
has been established in Eq.~\eqref{e:mu-ind.matrix.linear.solution},
one can obtain the final expression for the density matrix change 
that reproduces the induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
%\begin{multline}  \label{e:susceptibility-B} % PREPRINT
\begin{equation}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i}^{(0)} {\bf b}_{\beta}^{(i;1)} + C_{\beta i}^{(0)} {\bf b}_\alpha^{(i;1)} %\\ PREPRINT
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma}^{(0)} C_{\beta i}^{(0)} + D_{\beta\gamma}^{(0)} C_{\alpha i}^{(0)}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{equation}
%\end{multline} % PREPRINT
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ 
describes the local linear polarization affinity 
of the density matrix element 
associated with the $i$th LMO centroid as well as the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all the elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule. 
However, they are valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavor of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.

@@

\begin{acknowledgments}
This project is carried out under POLONEZ programme which has received funding from the European Union's
Horizon~2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement 
No.~665778. This project is funded by National Science Centre, Poland 
(registration No.~2016/23/P/ST4/01720) within the POLONEZ 3 fellowship. 
\end{acknowledgments}
